---
title: "AllOfUs Accrual Forecasting"
author: "Shawn Garbett, Jonathan Doreau, Cindy Chen, Keri Ann"
date: "3/4/2022"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(forecast)
library(tseries)
library(ggplot2)
library(dplyr)
library(lubridate)

accrual_raw <- read.csv("core_ptp_by_week.csv")

shutdown_start <- as.Date("2020-02-24")
shutdown_stop  <- as.Date("2021-03-08")

accrual <- accrual_raw %>%
           mutate(date = as.Date(
             parse_date_time(date_trunc, order="Ymd HMS"))) %>%
           group_by(date) %>%
           summarise(gain=sum(count, na.rm=TRUE)) %>%
           mutate(total = cumsum(gain)) %>%
           mutate(shutdown = date >= shutdown_start &
                             date <  shutdown_stop) %>%
           mutate(week = 1:n()) 
           
covid <- read.csv("United_States_COVID-19_Cases_and_Deaths_by_State_over_Time.csv") %>% 
  mutate(submission_date = as.Date(parse_date_time(submission_date, order="mdY"))) %>%
  mutate(date = as.Date(cut(submission_date, "week"))) %>%
  group_by(date) %>%
  summarise(covid_cases = sum(new_case), covid_death=sum(new_death)) 


accrual <- left_join(accrual, covid, by="date")
accrual$covid_cases[is.na(accrual$covid_cases)] <- 0
accrual$covid_death[is.na(accrual$covid_death)] <- 0

accrual$accel <- c(NA, diff(accrual$gain))
accrual$week  <- ts(accrual$week,  frequency=52)
accrual$gain  <- ts(accrual$gain,  frequency=52)
accrual$total <- ts(accrual$total, frequency=52)
accrual$accel <- ts(accrual$accel, frequency=52)

shutdown_pos_start <- match(TRUE, accrual$shutdown)
shutdown_pos_stop  <- nrow(accrual) - match(TRUE, rev(accrual$shutdown)) +1
```

## Problem Statement

Need: Forecast with confidence intervals for AllOfUs participation

Issues:

- COVID-19 shocks strongly affects accrual process.
- ARIMA forecasts from data are pessimistic and assume more shutdowns.

## Data

Data source is AllOfUs PDR

- Binned by Week
- Non-Test
- Non-Withdrawn
- Week is determined by completion of all core elements
  - Primary EHR Consent
  - PPI 1-3
  - Biosample suitable for DNA
  - Physical Measurements

## Raw Data Plot

```{r raw_data}
covid_rect <- function(y)
{
  geom_rect(aes(xmin=accrual$date[shutdown_pos_start],
                xmax=accrual$date[shutdown_pos_stop],
                ymin=min(y, na.rm=TRUE),
                ymax=max(y, na.rm=TRUE),
                fill='Shutdown'),
            alpha=0.5)
}

ggplot() + 
  geom_line(data = accrual, aes(x = date, y = as.numeric(gain))) + ylab('Weekly Gain') +
  covid_rect(accrual$gain)
```

Shutdown state will be used as covariate in prediction model.

## Covid Case Rate and Deaths

Additionally covariates for COVID-19 case rate and deaths would help inform the model.

The CDC provides public downloads of official case rates and deaths.

https://covid.cdc.gov/covid-data-tracker/#trends_dailycases

In a prediction model, the user will have to provide covid case/death forecasts. 

## Covid Plots

```{r plot_covid}

ggplot(covid, aes(date)) +
  geom_line(aes(y=covid_cases, colour="Cases"))+
  geom_line(aes(y=covid_death, colour="Deaths")) +
  ylab("Count") + 
  xlab("Week")

```

## Predictive Demostration

Leaving out the last year, how well does a model perform?

```{r predict_demo}

train   <- accrual[1:197,]
test    <- accrual[198:249,]

train$total <- ts(train$total, frequency=52)
test$total <- ts(test$total, frequency=52)

model <- auto.arima(train$total, seasonal=TRUE, allowdrift=TRUE, xreg=cbind(train$shutdown, train$covid_cases, train$covid_death))

fc <- forecast(model, h=52, xreg=cbind(rep(0, 52), test$covid_cases, test$covid_death))
plot(fc, xlab="Year", ylab="Total Participants",
     main="ARIMA Forecasting Last Year",
     sub="All Of Us Participants", xlim=c(4.5,6),ylim=c(2.5e5,3.5e5))
lines(198:249/52 + 1, test$total, lwd=2)

```

## Five Year Forecast

Forecasting forward requires assumptions about:

- Shutdown 
- COVID-19 Cases 
- COVID-19 Deaths

Our Assumptions:

- No more shutdowns
- 175000 weekly cases
- 6000 weekly deaths

## Five Year Forecast Results

```{r five_year}
model <- auto.arima(accrual$total, seasonal=TRUE, allowdrift=TRUE, 
                    xreg=cbind(accrual$shutdown,
                               accrual$covid_cases,
                               accrual$covid_death))
 
f_weeks <- 52*5
fc <- forecast(model, h=f_weeks, xreg=
                  cbind(rep(     0, f_weeks),
                        rep(175000, f_weeks),
                        rep(  6000, f_weeks)))
plot(fc, xlab="Year", ylab="Total Participants")
```

## Conclusions

- ARIMA forecasting needs assumptions about COVID.
- Confidence intervals are dependent on assumptions.
- Predictions intervals are reasonable.
- Prediction isn't perfect, but ARIMA is best known method.

## Appendix: Model Details

```{r mod_detail}
summary(model)
```

## Appendix: Model Diagnostics

```{r model_diag}
checkresiduals(model)
```


## Appendix: PDR Data Query

```
SELECT date_trunc('week',enrl_core_participant_time)
     , COUNT(participant_id)
     , site_id
     , hpo_id
     , b.organization_TYPE_ID

FROM pdr.mv_participant_all
LEFT JOIN mv_hpo b USING (hpo_id)
WHERE enrl_core_participant_time IS NOT NULL
AND test_participant = 0
GROUP BY 1,3,4,5
```
